---
title: "bee-percipitation-trends-scratch"
author: "jt-miller"
date: "2023-05-04"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Here is my scratch work for trying to see if percipitation trends in the deserts plays a role in diversity for bees

Hypothesis 1: The Two Distinct Rain Seasons facilitates increased diversity in the Sonoran Desert by opening up opportunity for different assemblages of (probably specialist) bees to inhabit this region.

Exploratory Steps: 

1. Identify sample grids of strong sampling for multiple years. How does sampling look across the year? 
2. Look at the bee diversity via occurrence records, check to see if the summer vs winter assemblage contains the same or different species.
      I. How Many species do these seasons contain? 
      II. How much overlap is there between the seasons in bee species? How many unique species
```{r}
# Load in the Sonoran Bees 
sonoran_bees <- read.delim("/blue/soltis/millerjared/pollen-project/Pollen-DB/processed-data/ecoregion-subset/bee_occurrence_data_Sonoran-Desert.txt")

# Evaluate Areas with Equal Sampling Effort 

```

