---
title: "percipitation-mapping"
author: "jt-miller"
date: "2023-04-24"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Percipitation Desert Mapping 

Libraries
```{r}
# Install the necessary packages
#remotes::install_github("mikejohnson51/AOI") # suggested!
#remotes::install_github("mikejohnson51/climateR")
# Call libraries 
library(AOI)
library(climateR)
library(sf)
library(tidyverse)
library(ggplot2)
library(raster)
library(terra)
library(tidyr)
library(tidyterra)
```

```{r}
# Read in shapefiles of the NA deserts of interest 

na_ecoregions_3 <- sf::read_sf("/blue/soltis/millerjared/Desert-Bees-Plants/raw-data/shapefiles/NA_CEC_Eco_Level3.shp")

# Subset out shapefiles to include only our deserts of interest 
na_deserts_3 <- subset(na_ecoregions_3, NA_L3NAME %in% c("Sonoran Desert", "Mojave Basin and Range")) # Maintain Native CRS

# Create grids for these ecoregions...
mojave <- subset(na_deserts_3, NA_L3NAME == "Mojave Basin and Range")
sonoran <- subset(na_deserts_3, NA_L3NAME == "Sonoran Desert" )


# Assign Cellside 
cellside <- 10000 # 10,000 m = 10km
mojave_grid_o <- sf::st_make_grid(mojave, 
                                cellsize = cellside) # cellsize = 10km x 10km 
mojave_grid <- sf::st_intersection(mojave_grid_o, mojave)

sonoran_grid_o <- sf::st_make_grid(sonoran, 
                                cellsize = cellside) # cellsize = 10km x 10km 
sonoran_grid <- sf::st_intersection(sonoran_grid_o, sonoran)

```

Plot these regions out 
```{r}
ggplot() + 
  geom_sf(sonoran_grid, mapping = aes()) + 
  geom_sf(mojave_grid, mapping = aes()) +
  theme_bw() +
  ggtitle("Mojave & Sonoran Ecoregions")
```
Create Custom Boundary Pull for Climate data
```{r}
# Using climateR for this analysis:
# Check the assets for relevant info 
p <- climateR::params # Pull out the relevant info for params 

p_tr <- filter(p, grepl("terraclim", id)) # Using TerraClim 
p_ppt <- filter(p_tr, varname == "ppt" ) # Relevant info for our precipitation pulls 

# Create AOIs for our pulls, define the crs to be the same as that of the percipitation data (unsure how necessary this is as it wasnt illustrated in the prior example but lets go with it)

# First transform coordinate reference system to match the TerraClim's projection 
sonoran_t <- st_transform(sonoran, crs = "+proj=longlat +a=6378137 +f=0.00335281066474748 +pm=0 +no_defs")
mojave_t <- st_transform(mojave, crs = "+proj=longlat +a=6378137 +f=0.00335281066474748 +pm=0 +no_defs")

# Assign AOI by screating an sfc object
sonoran_aoi <- st_as_sfc(sonoran_t) 
mojave_aoi <- st_as_sfc(mojave_t)

# Construct Raster Pulls usign TerraClim
sonoran_tc = getTerraClim(sonoran_aoi, varname = "ppt", startDate = "1990-01-01", endDate = "1990-12-31") # Monthly data ppt data
mojave_tc <- getTerraClim(mojave_aoi, varname = "ppt", startDate = "1990-01-01", endDate = "1990-12-31")

terra::res(sonoran_tc[[1]]) # 1/24th of a degree = 4x4km resolution

# Chirps is daily data, Consider using it later on? 
#chirps = getCHIRPS(na_deserts_3, startDate = "1990-01-01", endDate = "1990-12-01" )
```

Pull the Percipitation data out of the select Raster Layer
```{r}
# Traditional way to pull out raster data via dataframe 
tcdf <- terra::as.data.frame(tc[[1]], xy=TRUE, na.rm = TRUE) # Pull out the TerraClim Raster info 

```

Now we have xy coords which are 4x4 kms, We are looking to summarize these into our grid cells
```{r}
# Transform crs of the grids to match the crs of the raster data
sonoran_g_reproj <- st_transform(sonoran_grid, crs = "+proj=longlat +a=6378137 +f=0.00335281066474748 +pm=0 +no_defs")

mojave_g_reproj <- st_transform(mojave_grid, crs = "+proj=longlat +a=6378137 +f=0.00335281066474748 +pm=0 +no_defs")

# Assign IDs
sonoran_g_reproj <- sonoran_g_reproj %>%
  st_as_sf() %>% 
  mutate(ID = 1:n())
# Assign CRS 
tcdf_sf <- st_as_sf(tcdf, coords = c("x","y"), crs =  "+proj=longlat +a=6378137 +f=0.00335281066474748 +pm=0 +no_defs")

# Crop out the xy data by the sonoran region of interest
tcdf_cropped <- tcdf_sf[sonoran_aoi,]

# Group data into each of the grid cells

tcdf_gridded <- st_join(sonoran_g_reproj, tcdf_cropped) # Join together by spatial coords 
```

Another method as found on climateR's repo for extracting raster data 
```{r}
sonoran_dat = cbind(sonoran_g_reproj, exact_extract(sonoran_tc[[1]], sonoran_g_reproj, "mean", progress = FALSE)) # testing exact_extract as a function, supposed to take the mean of each extraction? 
mojave_dat = cbind(mojave_g_reproj, exact_extract(mojave_tc[[1]], mojave_g_reproj, "mean", progress = FALSE))
```



### Analysis (prototype)

First we want to take the avg percipitation per cell
```{r}
# Avg out precipitation across the grid cells. 
agg_pet_avg <- aggregate(tcdf_gridded[, 1:12], list(tcdf_gridded$ID), mean)

aggs <- st_drop_geometry(agg_pet_avg) # sf stuff messes with rowSums

# Create a new field looking at the total percipitation 
pet_total_df <- aggs%>% 
  #select(!'pet_1991-01-01 00:00:00_total') %>% 
  mutate(total_ppt = rowSums(.[2:13])) 
 # mutate(winter_pet = rowSums(.[]))
  
joined_data <- left_join(pet_total_df, agg_pet_avg[,c(1,14)], by = "Group.1") # Join the spatial data back up



joined_data_sf <- st_as_sf(joined_data)

# Of particular note, May has "0"s listed for many points. I presume that this is not actually 0, as 0.0 (floats) are used to denote this. Likely NA as in not measured? 

# Create an overall full percipitation
ggplot() + 
  geom_sf(joined_data_sf, mapping = aes(fill = total_ppt)) + 
  theme_bw() +
  ggtitle("Mojave percipitation totals")

# Create a timeseries 
# flip!
joined_data_piv <- joined_data %>% 
  pivot_longer(
    cols = contains("ppt_"),
    names_to = "ppt_month", 
    values_to = "ppt_avg"
   ) %>%
   mutate(ppt_month = case_when(
     grepl("1990-01-01", ppt_month) ~ "Jan",
     grepl("1990-01-31", ppt_month) ~ "Feb",# Closest we can get...
     grepl("1990-03-02", ppt_month) ~ "Mar",
     grepl("1990-04-02", ppt_month) ~ "Apr",
     grepl("1990-05-02", ppt_month) ~ "May",
     grepl("1990-06-02", ppt_month) ~ "Jun",
     grepl("1990-07-02", ppt_month) ~ "Jul",
     grepl("1990-08-01", ppt_month) ~ "Aug",
     grepl("1990-09-01", ppt_month) ~ "Sep",
     grepl("1990-10-01", ppt_month) ~ "Oct",
     grepl("1990-11-01", ppt_month) ~ "Nov",
     grepl("1990-12-01", ppt_month) ~ "Dec",


     TRUE ~ ppt_month
    ))

# Now calc the %percipitation per month to make our bar charts more informative
joined_data_piv <- joined_data_piv %>%
  group_by(Group.1, ppt_month) %>% 
  mutate(avg_month_ppt_percent = (ppt_avg/total_ppt)) %>% 
  mutate(avg_month_ppt_percent = round(avg_month_ppt_percent, digits = 4))

x <- c("Nov", "Dec", "Jan", "Feb", "Mar", "Apr", "May", "Jun", "Jul", "Aug", "Sep", "Oct")
 joined_data_piv$ppt_month = factor(joined_data_piv$ppt_month, levels = x)
 y <-  factor(joined_data_piv$ppt_month, levels = x)
 joined_data_piv_t <- joined_data_piv %>%
   filter(Group.1 == 1)
my_outlines <- c("darkblue", "darkblue", "darkblue", "darkblue", "darkblue",  "grey", "grey", "grey" , "darkorange", "darkorange", "darkorange" ,"grey")

scale_fill_ppt_months <- function(...){
    ggplot2:::manual_scale(
        'fill', 
        values = setNames(c("darkblue", "darkblue", "darkblue", "darkblue", "darkblue",  "grey", "grey", "grey" , "darkorange", "darkorange", "darkorange" ,"grey"), levels(y)), 
        ...
    )
}

scale_color_ppt_months <- function(...){
    ggplot2:::manual_scale(
        'color', 
        values = setNames(c("darkblue", "darkblue", "darkblue", "darkblue", "darkblue",  "grey", "grey", "grey" , "darkorange", "darkorange", "darkorange" ,"grey"), levels(y)), 
        ...
    )
}

names(my_outlines) <- levels(factor(levels(joined_data_piv$ppt_month)))
my_cols <- scale_fill_manual(name = "ppt_month", values = my_outlines)


ggplot(joined_data_piv_t, aes(x = ppt_month, y = ppt_avg,color = ppt_month)) +
  geom_bar(stat = "identity", fill = "lightblue") + 
  geom_text(aes( label = scales::percent(avg_month_ppt_percent),
                   y=ppt_avg), stat= "identity", vjust = -.5, show.legend = FALSE) +
  scale_color_ppt_months() + 
  ggtitle("Sample Site 10x10 km^2 1990 Percipitation Averaged by Month") +
  
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5)) +
  ylab("percipitation (mm)") +
  xlab("month") 


```
```{r}
# Testing with Extract...

# flip!
joined_data_piv_t <- dat %>% 
  pivot_longer(
    cols = contains("ppt_"),
    names_to = "ppt_month", 
    values_to = "ppt_avg"
   ) %>%
   mutate(ppt_month = case_when(
     grepl("1990.01.01", ppt_month) ~ "Jan",
     grepl("1990.01.31", ppt_month) ~ "Feb",# Closest we can get...
     grepl("1990.03.02", ppt_month) ~ "Mar",
     grepl("1990.04.02", ppt_month) ~ "Apr",
     grepl("1990.05.02", ppt_month) ~ "May",
     grepl("1990.06.02", ppt_month) ~ "Jun",
     grepl("1990.07.02", ppt_month) ~ "Jul",
     grepl("1990.08.01", ppt_month) ~ "Aug",
     grepl("1990.09.01", ppt_month) ~ "Sep",
     grepl("1990.10.01", ppt_month) ~ "Oct",
     grepl("1990.11.01", ppt_month) ~ "Nov",
     grepl("1990.12.01", ppt_month) ~ "Dec",


     TRUE ~ ppt_month
    )) %>% 
  filter(!ppt_month == "mean.ppt_1991.01.01.00.00.00_total") # Extraneous month grabbed 

joined_data_piv_t <- joined_data_piv_t %>%
  group_by(ID) %>% 
  mutate(total_ppt = sum(ppt_avg)) %>% 
  mutate(avg_month_ppt_percent = (ppt_avg/total_ppt)) %>% 
  mutate(avg_month_ppt_percent = round(avg_month_ppt_percent, digits = 4))


x <- c("Nov", "Dec", "Jan", "Feb", "Mar", "Apr", "May", "Jun", "Jul", "Aug", "Sep", "Oct")
 joined_data_piv_t$ppt_month = factor(joined_data_piv_t$ppt_month, levels = x)
 y <-  factor(joined_data_piv_t$ppt_month, levels = x)
names(my_outlines) <- levels(factor(levels(joined_data_piv_t$ppt_month)))
my_cols <- scale_fill_manual(name = "ppt_month", values = my_outlines)

joined_data_piv_t2 <- joined_data_piv_t[1:12,]

ggplot(joined_data_piv_t2, aes(x = ppt_month, y = ppt_avg,color = ppt_month)) +
  geom_bar(stat = "identity", fill = "lightblue") + 
  geom_text(aes( label = scales::percent(avg_month_ppt_percent),
                   y=ppt_avg), stat= "identity", vjust = -.5, show.legend = FALSE) +
  scale_color_ppt_months() + 
  ggtitle("Sample Site 10x10 km^2 1990 Percipitation Averaged by Month") +
  
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5)) +
  ylab("percipitation (mm)") +
  xlab("month") 



```
### Testing Grounds 
```{r}
# Finding the points according to the groupings...
test <- joined_data_piv_t %>% 
  filter(ID == "1")

# Plot them
ggplot() + 
  geom_sf(joined_data_piv_t, mapping = aes()) +
  geom_sf(test, mapping = aes(fill = "red")) # Yes appears that it does extract one grid cell at a time...

```


Further Summarize, create grid noting where the bimodal gradient occurs 
```{r}
# First create categories that define the winter & summer seasons for rainfall
summer_rains <- c("Jul", "Aug", "Sep")
winter_rains <- c("Nov", "Dec", "Jan", "Feb", "Mar")

joined_w_seasons <- joined_data_piv %>%
  mutate(rain_seasons = case_when(ppt_month %in% winter_rains ~ "winter_season",
                                 ppt_month %in% summer_rains ~ "summer_season",
                                TRUE ~ "else_months") )

# Sum up percentages of annual mean ppt by season for each gridcell
joined_w_season_o <- joined_w_seasons %>% 
   group_by(Group.1, rain_seasons) %>% 
   summarise(season_ppt_perc = sum(avg_month_ppt_percent))


```
# Look a this by exact_extract methods 
```{r}
summer_rains <- c("Jul", "Aug", "Sep")
winter_rains <- c("Nov", "Dec", "Jan", "Feb", "Mar")

joined_w_seasons_t <- joined_data_piv_t  %>%
  mutate(rain_seasons = case_when(ppt_month %in% winter_rains ~ "winter_season",
                                 ppt_month %in% summer_rains ~ "summer_season",
                                TRUE ~ "else_months") )

joined_w_season_o_t<- joined_w_seasons_t %>% 
   group_by(ID, rain_seasons) %>% 
   reframe(season_ppt_perc = sum(avg_month_ppt_percent), total_ppt)

joined_w_season_o_t2<- joined_w_seasons_t %>% 
   group_by(ID, rain_seasons) %>% 
   mutate(season_ppt_perc = sum(avg_month_ppt_percent))

son_spatial_summary_df <- distinct(joined_w_season_o_t2, ID, x, total_ppt, rain_seasons, season_ppt_perc)


son_summary_df <- distinct(joined_w_season_o_t, ID, rain_seasons, season_ppt_perc, total_ppt, .keep_all = TRUE)

```
### Create a plot to illustrate the distribution of rain % over the Deserts
```{r}
### Sonoran 

# Compare the % of Winter percipitation and % of summer percipitation as compared the average percipitation per grid cell. 

ggplot(son_summary_df, mapping = aes(x = total_ppt, y = season_ppt_perc)) +
  geom_point() + 
  facet_wrap(~rain_seasons) + 
  theme_bw() +
  ggtitle("Precipitation Relation by Season to 1990s averaged sum") + 
  xlab("Total Percipitation (mm)") +
  ylab("% Percipitation by Season")

## Justify the W -> E claim.

# First just by looking at general trends by overlaying on a map 
ggplot() + 
  geom_sf(son_spatial_summary_df, mapping = aes(fill = season_ppt_perc)) + 
  facet_wrap(~rain_seasons) +
  scale_fill_gradientn(
      colours = colorRampPalette((RColorBrewer::brewer.pal(11, "YlGnBu")))(9), 
      na.value = "#808080" # The line that denotes NAs as grey
    ) + 
  theme_bw()

# Convert geom to lon lat so we can do some scatterplotting 
centroid_son_summary_df <- son_spatial_summary_df %>% 
  st_centroid()


xy_df <- centroid_son_summary_df %>% 
    ungroup() %>% 
   dplyr::mutate(lon = sf::st_coordinates(.)[,1],
                lat = sf::st_coordinates(.)[,2])

ggplot(xy_df, aes(x = lon, y = season_ppt_perc, col = lat)) +
  geom_point() +
  scale_color_viridis_b() +
  facet_wrap(~rain_seasons) +
  ggtitle("Sonoran Seasonal Trends across Longitude") +
  xlab("Decimal Longitude") + 
  ylab("Percentage of Seasonal Percipitation") +
  theme_bw()

# ...

```



















# Join back up with the gridded map
```{r}

group_coords <- joined_w_seasons %>% 
  group_by(Group.1) %>% 
  distinct(rain_seasons, .keep_all = TRUE) %>% 
  ungroup() %>% 
  select(Group.1, geometry)
  
joined_w_season_o

joined_gridded <-left_join(group_coords, joined_w_season_o, by = "Group.1") # Unsure whats happening with the duplication here? Possibly geometry messing things up 

joined_gridded <- joined_gridded %>% 
  distinct(Group.1, geometry, rain_seasons, season_ppt_perc, .keep_all = TRUE ) # Distinct on all columns, removing true duplicates

joined_gridded <- st_as_sf(joined_gridded, crs = "+proj=longlat +a=6378137 +f=0.00335281066474748 +pm=0 +no_defs" )
```

# Create a plot! 
```{r}
ggplot(joined_gridded) +
    geom_sf(aes(fill = (season_ppt_perc))) + # Log Transform, correct for log10(1) = 0
    theme_bw() +
    scale_fill_gradientn(
      colours = colorRampPalette((RColorBrewer::brewer.pal(11, "YlOrRd")))(9), 
      na.value = "#808080" # The line that denotes NAs as grey
    ) +
    
    #ggtitle(paste0("Bee Sampling Intensity 10x10km Grid", "\n", region_name_v[[i]])) +
    theme(axis.text.x = element_text(angle = 45, vjust = 0.5)) +
    #ggtitle("Bee Sampling Intensity 10x10km Grid") +
    theme(plot.title = element_text(hjust = 0.5)) +
    guides(fill=guide_legend(title="Log10 Occurrences"))
```





Plot
```{r}
library(patchwork)

ggplot() +
  geom_spatraster(data = tc$pet) +
  facet_wrap(~lyr) +
  scale_fill_whitebox_c(
    palette = "muted",
    na.value = "white"
  ) + 
  geom_sf(data = sonoran_aoi, fill = NA, lwd = 1, col = "black") + 
  theme_minimal() 


ggplot() + 
  geom_sf(tcdf_sf, mapping = aes()) # Doesnt actually crop...

ggplot() + 
  geom_sf(tcdf_cropped, mapping = aes())
```


